---
title: "FINAL_1"
author: "Najeeba Shaikh"
date: "2025-07-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("readr")
library("dplyr")
library("ggplot2")
library("sf")
library("tidyr")
library("mapview")
library("viridis")
library("viridisLite")
library("vegan")

ncld_detailed_percentage <- read.csv("~/Desktop/PINC_Climate_Change/Project 2/nlcd_2001_detailed_percentages.csv")

colnames(ncld_detailed_percentage) <- c(
  "grid_cell",                         # grid_cell (assuming it's the first column)
  "Open_Water",                        # 11
  "Developed_Open_Space",              # 21
  "Developed_Low_Intensity",           # 22
  "Developed_Medium_Intensity",        # 23
  "Developed_High_Intensity",          # 24
  "Barren_Land_Rock_Sand_Clay",        # 31
  "Deciduous_Forest",                  # 41
  "Evergreen_Forest",                  # 42
  "Mixed_Forest",                      # 43
  "Shrub_Scrub",                       # 52
  "Grassland_Herbaceous",              # 71
  "Pasture_Hay",                       # 81
  "Cultivated_Crops",                  # 82
  "Woody_Wetlands",                    # 90
  "Emergent_Herbaceous_Wetlands"       # 95
)

read.csv("~/Desktop/PINC_Climate_Change/Project 2/california_bioclim.csv")
bio_clim <- read.csv("~/Desktop/PINC_Climate_Change/Project 2/california_bioclim.csv")

colnames(bio_clim) <- c(
  "grid_cell",                          # grid_cell (assuming it's the first column)
  "Annual_Mean_Temperature",            # BIO1
  "Mean_Diurnal_Range",                 # BIO2
  "Isothermality",                      # BIO3
  "Temperature_Seasonality",            # BIO4
  "Max_Temperature_of_Warmest_Month",   # BIO5
  "Min_Temperature_of_Coldest_Month",   # BIO6
  "Temperature_Annual_Range",           # BIO7
  "Mean_Temperature_of_Wettest_Quarter", # BIO8
  "Mean_Temperature_of_Driest_Quarter", # BIO9
  "Mean_Temperature_of_Warmest_Quarter",# BIO10
  "Mean_Temperature_of_Coldest_Quarter",# BIO11
  "Annual_Precipitation",               # BIO12
  "Precipitation_of_Wettest_Month",     # BIO13
  "Precipitation_of_Driest_Month",      # BIO14
  "Precipitation_Seasonality",          # BIO15
  "Precipitation_of_Wettest_Quarter",   # BIO16
  "Precipitation_of_Driest_Quarter",    # BIO17
  "Precipitation_of_Warmest_Quarter",   # BIO18
  "Precipitation_of_Coldest_Quarter"    # BIO19
)

processed_diversity <- read.csv("~/Desktop/PINC_Climate_Change/Project 2/processed_diversity.csv")

# merge the datasets
DATA_1 <- merge(bio_clim, ncld_detailed_percentage, by = "grid_cell")

DATA <- merge(DATA_1, processed_diversity, by = "grid_cell")

#make graph
mapview(DATA, xcol = "X_Axis_0_", ycol = "X_Axis_1_", crs = 3310, grid = TRUE)

# Convert DATA to spatial (sf) object
DATA_sf <- st_as_sf(DATA, coords = c("X_Axis_0_", "X_Axis_1_"), crs = 3310)
mapview(DATA_sf, zcol = "Developed_High_Intensity")

ggplot(DATA_sf) +
  geom_sf(aes(color = Developed_Open_Space), size = 1) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "Developed Open Space in California")

DATA_sf$Total_Developed <- DATA_sf$Developed_Open_Space +
  DATA_sf$Developed_Low_Intensity +
  DATA_sf$Developed_Medium_Intensity +
  DATA_sf$Developed_High_Intensity

ggplot(DATA_sf) +
  geom_sf(aes(color = Total_Developed), size = 1) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "Total Developed Land in California")

mapview(DATA_sf, zcol = "Developed_Low_Intensity")
mapview(DATA_sf, zcol = "Developed_Medium_Intensity")
mapview(DATA_sf, zcol = "Developed_High_Intensity")

library(tidyr)
library(RColorBrewer)

# Pivot longer for faceted plotting
DATA_long <- DATA_sf %>%
  pivot_longer(cols = c(Developed_Low_Intensity,
                        Developed_Medium_Intensity,
                        Developed_High_Intensity),
               names_to = "Development_Type",
               values_to = "Percentage")

ggplot(DATA_long) +
  geom_sf(aes(color = Percentage), size = 0.8) +
  scale_color_viridis_c(option = "C") +
  facet_wrap(~ Development_Type) +
  theme_minimal() +
  labs(title = "Developed Land Intensity Types in California")

plot(st_geometry(DATA_sf))  # Should look like California!


```
```{r}
ggplot(DATA_long) +
  geom_sf(aes(color = Percentage), size = 0.8) +
  scale_color_viridis_c(option = "C") +
  facet_wrap(~ Development_Type, ncol = 1, strip.position = "top") +  # One column, clearer labels
  theme_minimal(base_size = 14) +  # Larger base font
  theme(
    strip.text = element_text(size = 14, face = "bold"),  # Facet labels
    axis.text.x = element_text(angle = 45, hjust = 1),     # Tilt x-axis labels
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  ) +
  labs(
    title = "Developed Land Intensity Types in California",
    color = "Percent Cover"
  )

```

```{r}
# Clean labels before plotting
DATA_long$Development_Type <- recode(DATA_long$Development_Type,
  Developed_Low_Intensity = "Low Intensity",
  Developed_Medium_Intensity = "Medium Intensity",
  Developed_High_Intensity = "High Intensity"
)

# Plot with horizontal layout and clean labels
ggplot(DATA_long) +
  geom_sf(aes(color = Percentage), size = 0.8) +
  scale_color_viridis_c(option = "C") +
  facet_wrap(~ Development_Type, nrow = 1) +  # horizontal layout
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    axis.text.x = element_blank(),   # remove x-axis labels
    axis.ticks.x = element_blank(),
    panel.grid = element_blank()
  ) +
  labs(
    title = "Developed Land Intensity Types in California",
    color = "Percentage"
  )

```

```{r}
library(ggplot2)
library(tidyr)

DATA_long <- DATA_sf %>%
  pivot_longer(cols = c(Developed_Open_Space,
                        Developed_Low_Intensity,
                        Developed_Medium_Intensity,
                        Developed_High_Intensity),
               names_to = "Development_Type",
               values_to = "Percentage")

ggplot(DATA_long, aes(x = Development_Type, y = Percentage, fill = Development_Type)) +
  geom_boxplot() +
  scale_fill_viridis_d(option = "C") +
  theme_minimal(base_size = 14) +
  labs(title = "Boxplot of Developed Land Types",
       x = "Development Type",
       y = "Percentage") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
library(dplyr)

summary_data <- DATA_long %>%
  group_by(Development_Type) %>%
  summarise(mean_pct = mean(Percentage),
            sd_pct = sd(Percentage),
            .groups = "drop")

ggplot(summary_data, aes(x = Development_Type, y = mean_pct, fill = Development_Type)) +
  geom_bar(stat = "identity", color = "black") +
  geom_errorbar(aes(ymin = mean_pct - sd_pct, ymax = mean_pct + sd_pct), width = 0.2) +
  scale_fill_viridis_d(option = "C") +
  theme_minimal(base_size = 14) +
  labs(title = "Mean Percentage of Development Types",
       x = "Development Type",
       y = "Mean Percentage") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
library(dplyr)

binned_data <- DATA_long %>%
  mutate(Percent_Bin = cut(Percentage, breaks = seq(0, 100, by = 10))) %>%
  group_by(Development_Type, Percent_Bin) %>%
  summarise(mean_richness = mean(X_ENDC_RICHNESS_, na.rm = TRUE),
            .groups = "drop")

ggplot(binned_data, aes(x = Percent_Bin, y = mean_richness, color = Development_Type, group = Development_Type)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_color_viridis_d(option = "D") +
  theme_minimal(base_size = 14) +
  labs(title = "Mean Species Richness by Binned Land Cover Intensity",
       x = "Binned % Land Cover",
       y = "Mean Species Richness") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
# High and medium intensity development reduce biodiversity when they dominate a grid cell.
# Low-intensity development has a hump-shaped relationship, possibly peaking in ecotones or transitional zones.
# The surprising peak in high-intensity (20–30%) might be due to:
# A small number of urban-edge sites that still support biodiversity.
# Averaging across few data points, leading to skew.





```{r}
heat_data <- DATA_long %>%
  mutate(Percent_Bin = cut(Percentage, breaks = seq(0, 100, by = 5)),
         Richness_Bin = cut(X_ENDC_RICHNESS_, breaks = seq(floor(min(X_ENDC_RICHNESS_, na.rm = TRUE)),
                                                           ceiling(max(X_ENDC_RICHNESS_, na.rm = TRUE)), by = 2))) %>%
  group_by(Development_Type, Percent_Bin, Richness_Bin) %>%
  summarise(count = n(), .groups = "drop")

ggplot(heat_data, aes(x = Percent_Bin, y = Richness_Bin, fill = count)) +
  geom_tile() +
  facet_wrap(~ Development_Type) +
  scale_fill_viridis_c(option = "C", na.value = "white") +
  theme_minimal(base_size = 13) +
  labs(title = "Heatmap of Species Richness vs. Developed Land Cover",
       x = "% Land Cover (Binned)",
       y = "Species Richness (Binned)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


library(ggplot2)
library(dplyr)
library(tidyr)

# Re-bin with fewer breaks for readability
DATA_long <- DATA_sf %>%
  pivot_longer(cols = c(Developed_Open_Space,
                        Developed_Low_Intensity,
                        Developed_Medium_Intensity,
                        Developed_High_Intensity),
               names_to = "Development_Type",
               values_to = "Percentage") %>%
  mutate(
    Percent_Bin = cut(Percentage, breaks = seq(0, 100, by = 10), include.lowest = TRUE),
    Richness_Bin = cut(X_ENDC_RICHNESS_, breaks = 8)  # Fewer bins, auto-calculated range
  )

# Count observations in each bin
heat_data <- DATA_long %>%
  group_by(Development_Type, Percent_Bin, Richness_Bin) %>%
  summarise(count = n(), .groups = "drop")

# Improved heatmap
ggplot(heat_data, aes(x = Percent_Bin, y = Richness_Bin, fill = count)) +
  geom_tile(color = "white", linewidth = 0.1) +
  facet_wrap(~ Development_Type) +
  scale_fill_viridis_c(option = "D", direction = -1, na.value = "grey90") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Improved Heatmap: Richness vs. Land Development",
    x = "% Land Cover (Binned)",
    y = "Species Richness (Binned)",
    fill = "Count"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

```
# Low- and open-intensity development can coexist with biodiversity, but the relationship is variable.
# Medium- and high-intensity development are consistently associated with low richness, even at low percentages — a strong sign of ecological impact.
# Spatial planning could prioritize open space and limit medium/high intensity development in areas of high conservation value.

```{r}
library(ggplot2)
library(tidyr)

# Reshape development types into long format
DATA_long <- DATA_sf %>%
  pivot_longer(cols = c(Developed_Open_Space,
                        Developed_Low_Intensity,
                        Developed_Medium_Intensity,
                        Developed_High_Intensity),
               names_to = "Development_Type",
               values_to = "Percentage")

# Scatterplot with trend line
ggplot(DATA_long, aes(x = Percentage, y = X_ENDC_RICHNESS_)) +
  geom_point(alpha = 0.5, color = "black") +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  facet_wrap(~ Development_Type, scales = "free_x") +
  theme_minimal(base_size = 14) +
  labs(title = "Species Richness vs. Developed Land Types",
       x = "Percent Land Cover",
       y = "Species Richness")
```


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Development columns to reshape
cols_to_use <- c("Developed_Open_Space",
                 "Developed_Low_Intensity",
                 "Developed_Medium_Intensity",
                 "Developed_High_Intensity")

# Ensure columns are numeric
DATA_sf <- DATA_sf %>%
  mutate(across(all_of(cols_to_use), ~ suppressWarnings(as.numeric(as.character(.)))))

# Filter out rows where all dev columns are missing or invalid
DATA_filtered <- DATA_sf %>%
  filter(if_any(all_of(cols_to_use), ~ !is.na(.) & . >= 0 & . <= 100))

# Pivot to long format
DATA_long_full <- DATA_filtered %>%
  pivot_longer(cols = all_of(cols_to_use),
               names_to = "Development_Type",
               values_to = "Percentage")

# Final plot
ggplot(DATA_long_full, aes(x = Percentage, y = X_ENDC_RICHNESS_)) +
  geom_point(alpha = 0.5, color = "skyblue") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  facet_wrap(~ Development_Type, scales = "free_x") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Species Richness vs. Developed Land Types",
    x = "Percent Land Cover",
    y = "Species Richness"
  )

# add r^2 values and slopes

```


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)

# Define columns of interest
cols_to_use <- c("Developed_Open_Space",
                 "Developed_Low_Intensity",
                 "Developed_Medium_Intensity",
                 "Developed_High_Intensity")

# Ensure numeric
DATA_sf <- DATA_sf %>%
  mutate(across(all_of(cols_to_use), ~ suppressWarnings(as.numeric(as.character(.)))))

# Filter
DATA_filtered <- DATA_sf %>%
  filter(if_any(all_of(cols_to_use), ~ !is.na(.) & . >= 0 & . <= 100))

# Pivot to long
DATA_long <- DATA_filtered %>%
  pivot_longer(cols = all_of(cols_to_use),
               names_to = "Development_Type",
               values_to = "Percentage")

# Fit models per Development_Type
model_stats <- DATA_long %>%
  group_by(Development_Type) %>%
  summarise(
    slope = coef(lm(X_ENDC_RICHNESS_ ~ Percentage))[2],
    r_squared = summary(lm(X_ENDC_RICHNESS_ ~ Percentage))$r.squared,
    correlation = cor(Percentage, X_ENDC_RICHNESS_, use = "complete.obs", method = "pearson"),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("Slope = ", round(slope, 2),
                   "\nR² = ", round(r_squared, 2),
                   "\nr = ", round(correlation, 2))
  )

# Drop geometry to compute label positions safely
label_positions <- DATA_long %>%
  st_drop_geometry() %>%
  group_by(Development_Type) %>%
  summarise(
    x = max(Percentage, na.rm = TRUE) * 0.7,
    y = max(X_ENDC_RICHNESS_, na.rm = TRUE) * 0.9,
    .groups = "drop"
  ) %>%
  left_join(model_stats, by = "Development_Type")

# Final plot
ggplot(DATA_long, aes(x = Percentage, y = X_ENDC_RICHNESS_)) +
  geom_point(alpha = 0.5, color = "skyblue") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  facet_wrap(~ Development_Type, scales = "free_x") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Species Richness vs. Developed Land Types",
    x = "Percent Land Cover",
    y = "Species Richness"
  ) +
  geom_text(data = label_positions,
            aes(x = x, y = y, label = label),
            inherit.aes = FALSE,
            size = 3,  # Reduced text size
            hjust = 0)

```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)

# Define columns of interest
cols_to_use <- c("Developed_Open_Space",
                 "Developed_Low_Intensity",
                 "Developed_Medium_Intensity",
                 "Developed_High_Intensity")

# Ensure numeric
DATA_sf <- DATA_sf %>%
  mutate(across(all_of(cols_to_use), ~ suppressWarnings(as.numeric(as.character(.)))))

# Filter for reasonable percentages
DATA_filtered <- DATA_sf %>%
  filter(if_any(all_of(cols_to_use), ~ !is.na(.) & . >= 0 & . <= 100))

# Pivot to long
DATA_long <- DATA_filtered %>%
  pivot_longer(cols = all_of(cols_to_use),
               names_to = "Development_Type",
               values_to = "Percentage")

# Fit models per Development_Type
model_stats <- DATA_long %>%
  group_by(Development_Type) %>%
  summarise(
    slope = coef(lm(X_ENDC_RICHNESS_ ~ Percentage))[2],
    r_squared = summary(lm(X_ENDC_RICHNESS_ ~ Percentage))$r.squared,
    correlation = cor(Percentage, X_ENDC_RICHNESS_, use = "complete.obs"),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("Slope = ", round(slope, 2),
                   "\nR² = ", round(r_squared, 2),
                   "\nr = ", round(correlation, 2))
  )

# Drop geometry to position labels
label_positions <- DATA_long %>%
  st_drop_geometry() %>%
  group_by(Development_Type) %>%
  summarise(
    x = min(Percentage, na.rm = TRUE) + 1,
    y = max(X_ENDC_RICHNESS_, na.rm = TRUE) * 0.95,
    .groups = "drop"
  ) %>%
  left_join(model_stats, by = "Development_Type")

# Final plot
ggplot(DATA_long, aes(x = Percentage, y = X_ENDC_RICHNESS_)) +
  geom_point(alpha = 0.5, color = "skyblue") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  facet_wrap(~ Development_Type, scales = "free_x") +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(hjust = 0),  # left-align facet labels
    plot.title = element_text(size = 18, face = "bold")
  ) +
  labs(
    title = "Species Richness vs. Developed Land Types",
    x = "Percent Land Cover",
    y = "Species Richness"
  ) +
  geom_text(data = label_positions,
            aes(x = x, y = y, label = label),
            inherit.aes = FALSE,
            size = 2.8, hjust = 0, vjust = 1)

```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)

# Choose valid columns
cols_to_use <- c("Developed_Low_Intensity",
                 "Developed_Medium_Intensity",
                 "Developed_High_Intensity")

# Clean and filter
DATA_filtered <- DATA_sf %>%
  mutate(across(all_of(cols_to_use), ~ suppressWarnings(as.numeric(as.character(.))))) %>%
  filter(across(all_of(cols_to_use), ~ !is.na(.) & . >= 0 & . <= 100))

# Pivot to long
DATA_long <- DATA_filtered %>%
  pivot_longer(cols = all_of(cols_to_use),
               names_to = "Development_Type",
               values_to = "Percentage")

# Create facet-friendly labels
DATA_long$Development_Type <- factor(DATA_long$Development_Type,
                                     levels = cols_to_use,
                                     labels = c("Low Intensity", "Medium Intensity", "High Intensity"))

# Stats per type
model_stats <- DATA_long %>%
  group_by(Development_Type) %>%
  summarise(
    slope = coef(lm(X_ENDC_RICHNESS_ ~ Percentage))[2],
    r_squared = summary(lm(X_ENDC_RICHNESS_ ~ Percentage))$r.squared,
    correlation = cor(Percentage, X_ENDC_RICHNESS_, use = "complete.obs"),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("Slope = ", round(slope, 2),
                   "\nR² = ", round(r_squared, 2),
                   "\nr = ", round(correlation, 2))
  )

# Label positions
label_positions <- DATA_long %>%
  st_drop_geometry() %>%
  group_by(Development_Type) %>%
  summarise(
    x = min(Percentage, na.rm = TRUE) + 1,
    y = max(X_ENDC_RICHNESS_, na.rm = TRUE) * 0.95,
    .groups = "drop"
  ) %>%
  left_join(model_stats, by = "Development_Type")

# Custom colors
my_colors <- c("Low Intensity" = "pink", 
               "Medium Intensity" = "lavender", 
               "High Intensity" = "coral1")

# ✨ Final Poster-Worthy Plot
ggplot(DATA_long, aes(x = Percentage, y = X_ENDC_RICHNESS_, color = Development_Type)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, size = 1.2) +
  facet_wrap(~ Development_Type, scales = "free_x") +
  scale_color_manual(values = my_colors) +
  geom_text(data = label_positions,
            aes(x = x, y = y, label = label, color = Development_Type),
            inherit.aes = FALSE,
            size = 3.5, hjust = 0, fontface = "bold") +
  theme_minimal(base_size = 16) +
  theme(
    strip.text = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.title = element_text(size = 20, face = "bold"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 12),
    legend.position = "none"
  ) +
  labs(
    title = "Species Richness vs. Developed Land Cover Types",
    x = "Percent Land Cover",
    y = "Species Richness"
  )

```


```{r}
bee_matrix <- processed_diversity %>%
  select(grid_cell, starts_with("sp")) %>%
  mutate(across(-grid_cell, ~ replace_na(., 0)))

# Quick diagnostics
summary(rowSums(bee_matrix[-1]))
length(unique(as.data.frame(bee_matrix[-1])))

set.seed(123)

bee_matrix <- data.frame(grid_cell = paste0("cell_", 1:50))
for (i in 1:20) {
  bee_matrix[[paste0("sp", i)]] <- rpois(50, lambda = sample(0:5, 1))
}

# Remove empty or duplicate rows
bee_matrix <- bee_matrix %>%
  filter(rowSums(across(-grid_cell)) > 0) %>%
  distinct(across(-grid_cell), .keep_all = TRUE)

rownames(bee_matrix) <- bee_matrix$grid_cell
species_only <- bee_matrix %>% select(-grid_cell)

# Run NMDS
nmds_result <- metaMDS(species_only, k = 2, trymax = 100)


```

```{r}
library(ggplot2)
library(dplyr)
library(viridis)

# Step 1: Get NMDS coordinates
nmds_scores <- as.data.frame(nmds_result$points)
nmds_scores$grid_cell <- rownames(nmds_scores)

# Step 2: Add fake development % and bin it for color
set.seed(42)
nmds_scores$Total_Developed <- sample(0:100, size = nrow(nmds_scores), replace = TRUE)

nmds_scores <- nmds_scores %>%
  mutate(
    Dev_Category = cut(Total_Developed,
                       breaks = c(-1, 10, 25, 50, 100),
                       labels = c("Very Low", "Low", "Moderate", "High"))
  )

# Step 3: Plot the NMDS
ggplot(nmds_scores, aes(x = MDS1, y = MDS2, color = Dev_Category)) +
  geom_point(size = 3.5, alpha = 0.85) +
  scale_color_viridis_d(option = "C") +
  theme_minimal(base_size = 16) +
  labs(
    title = "NMDS of Simulated Bee Communities",
    subtitle = "Colored by Simulated Development Level",
    x = "NMDS1", y = "NMDS2", color = "Development"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

```

```{r}
library(ggplot2)
library(ggforce)    # for ellipses
library(viridis)
library(dplyr)

# Use your working NMDS scores with Dev_Category
ggplot(nmds_scores, aes(x = MDS1, y = MDS2, color = Dev_Category)) +
  # Ellipses to group development levels
  ggforce::geom_mark_ellipse(aes(label = Dev_Category, group = Dev_Category),
                             label.fontsize = 12, label.fill = "white",
                             color = "black", alpha = 0.05, expand = unit(1, "mm")) +
  
  # Bee points
  geom_point(size = 4, alpha = 0.9) +
  
  # Color palette
  scale_color_viridis_d(option = "C", begin = 0.2, end = 0.9) +
  
  # Labels and styling
  labs(
    title = "🧭 NMDS of Simulated Bee Communities",
    subtitle = "Grouped by Development Level",
    x = "NMDS1",
    y = "NMDS2",
    color = "Development"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 14),
    legend.position = "right",
    panel.grid = element_blank()
  )

```

```{r}
library(dplyr)
library(tidyr)
library(plotly)

# Clean and prepare your data
DATA_long <- DATA_sf %>%
  mutate(across(c(Developed_Open_Space,
                  Developed_Low_Intensity,
                  Developed_Medium_Intensity,
                  Developed_High_Intensity,
                  X_ENDC_RICHNESS_), ~ as.numeric(.))) %>%
  pivot_longer(cols = c(Developed_Open_Space,
                        Developed_Low_Intensity,
                        Developed_Medium_Intensity,
                        Developed_High_Intensity),
               names_to = "Development_Type",
               values_to = "Percent_Cover") %>%
  filter(!is.na(Percent_Cover) & Percent_Cover >= 0 & Percent_Cover <= 100) %>%
  mutate(
    Development_Type = recode(Development_Type,
                              Developed_Open_Space = "Open Space",
                              Developed_Low_Intensity = "Low Intensity",
                              Developed_Medium_Intensity = "Medium Intensity",
                              Developed_High_Intensity = "High Intensity")
  )

# Create one plot for each development type
plots <- DATA_long %>%
  group_split(Development_Type) %>%
  lapply(function(df) {
    plot_ly(df,
            x = ~Percent_Cover,
            y = ~X_ENDC_RICHNESS_,
            type = 'scatter',
            mode = 'markers',
            marker = list(size = 5, opacity = 0.7),
            name = unique(df$Development_Type),
            text = ~paste0("Richness: ", X_ENDC_RICHNESS_,
                           "<br>Percent: ", round(Percent_Cover, 1), "%"),
            hoverinfo = "text") %>%
      layout(
        title = list(text = unique(df$Development_Type), x = 0.5),
        xaxis = list(title = "% Land Cover"),
        yaxis = list(title = "Bee Species Richness")
      )
  })

# Combine all plots into subplots (2 rows × 2 columns)
subplot(plots, nrows = 2, shareX = TRUE, shareY = TRUE, titleX = TRUE, titleY = TRUE) %>%
  layout(title = "Interactive Bee Richness vs. Land Development Intensity",
         showlegend = FALSE)

```

```{r}
library(dplyr)
library(tidyr)
library(plotly)

# Prepare long-format data
DATA_long <- DATA_sf %>%
  mutate(across(c(Developed_Open_Space,
                  Developed_Low_Intensity,
                  Developed_Medium_Intensity,
                  Developed_High_Intensity,
                  X_ENDC_RICHNESS_), ~ as.numeric(.))) %>%
  pivot_longer(cols = c(Developed_Open_Space,
                        Developed_Low_Intensity,
                        Developed_Medium_Intensity,
                        Developed_High_Intensity),
               names_to = "Development_Type",
               values_to = "Percent_Cover") %>%
  filter(!is.na(Percent_Cover) & Percent_Cover >= 0 & Percent_Cover <= 100) %>%
  mutate(
    Development_Type = recode(Development_Type,
      Developed_Open_Space = "Open Space",
      Developed_Low_Intensity = "Low Intensity",
      Developed_Medium_Intensity = "Medium Intensity",
      Developed_High_Intensity = "High Intensity")
  )

# Unified interactive scatter plot
plot_ly(
  data = DATA_long,
  x = ~Percent_Cover,
  y = ~X_ENDC_RICHNESS_,
  color = ~Development_Type,
  colors = "Set2",
  type = "scatter",
  mode = "markers",
  marker = list(size = 5, opacity = 0.7),
  text = ~paste("Development:", Development_Type,
                "<br>Land Cover:", round(Percent_Cover, 1), "%",
                "<br>Richness:", X_ENDC_RICHNESS_),
  hoverinfo = "text"
) %>%
  layout(
    title = "Bee Species Richness vs. Land Development Intensity",
    xaxis = list(title = "% Land Cover"),
    yaxis = list(title = "Bee Species Richness"),
    legend = list(title = list(text = "<b>Development Type</b>"))
  )

```

```{r}
library(dplyr)
library(tidyr)
library(plotly)

# Step 1: Prepare data
DATA_long <- DATA_sf %>%
  mutate(across(c(Developed_Open_Space,
                  Developed_Low_Intensity,
                  Developed_Medium_Intensity,
                  Developed_High_Intensity,
                  X_ENDC_RICHNESS_), as.numeric)) %>%
  
  pivot_longer(cols = starts_with("Developed_"),
               names_to = "Development_Type",
               values_to = "Percent_Cover") %>%
  
  filter(!is.na(Percent_Cover), Percent_Cover >= 0, Percent_Cover <= 100) %>%
  
  # Force character matching to prevent factor duplicates
  mutate(
    Development_Type = as.character(Development_Type),
    Development_Type = case_when(
      Development_Type == "Developed_Open_Space" ~ "Open Space",
      Development_Type == "Developed_Low_Intensity" ~ "Low Intensity",
      Development_Type == "Developed_Medium_Intensity" ~ "Medium Intensity",
      Development_Type == "Developed_High_Intensity" ~ "High Intensity",
      TRUE ~ Development_Type
    ),
    Development_Type = factor(Development_Type, levels = c("Open Space", "Low Intensity", "Medium Intensity", "High Intensity"))
  )

# Step 2: Plot
plot_ly(
  data = DATA_long,
  x = ~Percent_Cover + 0.1,  # offset to allow log scale (no 0s)
  y = ~X_ENDC_RICHNESS_,
  color = ~Development_Type,
  colors = "Set2",
  type = 'scatter',
  mode = 'markers',
  marker = list(size = 5, opacity = 0.5),
  text = ~paste0("<b>", Development_Type, "</b><br>",
                 "Land Cover: ", round(Percent_Cover, 2), "%<br>",
                 "Richness: ", X_ENDC_RICHNESS_),
  hoverinfo = "text"
) %>%
  layout(
    title = "Bee Species Richness vs. Land Development Intensity",
    xaxis = list(title = "% Land Cover (log-scaled)", type = "log"),
    yaxis = list(title = "Bee Species Richness"),
    legend = list(title = list(text = "<b>Development Type</b>")),
    margin = list(l = 60, r = 40, t = 60, b = 50)
  )

```

```{r}
library(dplyr)
library(tidyr)
library(plotly)

# Step 1: Clean data
DATA_long <- DATA_sf %>%
  mutate(across(c(Developed_Open_Space,
                  Developed_Low_Intensity,
                  Developed_Medium_Intensity,
                  Developed_High_Intensity,
                  X_ENDC_RICHNESS_), as.numeric)) %>%
  pivot_longer(cols = starts_with("Developed_"),
               names_to = "Development_Type",
               values_to = "Percent_Cover") %>%
  filter(!is.na(Percent_Cover), Percent_Cover >= 0, Percent_Cover <= 100) %>%
  mutate(
    Development_Type = recode(Development_Type,
      "Developed_Open_Space" = "Open Space",
      "Developed_Low_Intensity" = "Low Intensity",
      "Developed_Medium_Intensity" = "Medium Intensity",
      "Developed_High_Intensity" = "High Intensity"
    )
  )

# Step 2: Build plot traces manually
dev_levels <- unique(DATA_long$Development_Type)
colors <- RColorBrewer::brewer.pal(length(dev_levels), "Set2")
plot <- plot_ly()

for (i in seq_along(dev_levels)) {
  dev_type <- dev_levels[i]
  df_sub <- DATA_long %>% filter(Development_Type == dev_type)

  # Linear model for trend line and R²
  fit <- lm(X_ENDC_RICHNESS_ ~ Percent_Cover, data = df_sub)
  r2 <- round(summary(fit)$r.squared, 2)
  line_df <- data.frame(
    Percent_Cover = seq(min(df_sub$Percent_Cover), max(df_sub$Percent_Cover), length.out = 100)
  )
  line_df$Predicted <- predict(fit, newdata = line_df)

  # Add points
  plot <- plot %>%
    add_trace(
      data = df_sub,
      x = ~Percent_Cover + 0.1,
      y = ~X_ENDC_RICHNESS_,
      type = 'scatter',
      mode = 'markers',
      name = paste0(dev_type, " (R² = ", r2, ")"),
      marker = list(size = 5, opacity = 0.5, color = colors[i]),
      showlegend = TRUE
    ) %>%
    # Add regression line
    add_trace(
      data = line_df,
      x = ~Percent_Cover + 0.1,
      y = ~Predicted,
      type = 'scatter',
      mode = 'lines',
      line = list(color = colors[i], width = 2),
      name = paste0(dev_type, " Trend"),
      showlegend = FALSE
    )
}

# Step 3: Layout
plot %>%
  layout(
    title = "Bee Species Richness vs. Land Development Intensity (with R²)",
    xaxis = list(title = "% Land Cover (log-scaled)", type = "log"),
    yaxis = list(title = "Bee Species Richness"),
    legend = list(title = list(text = "<b>Development Type</b>")),
    margin = list(l = 60, r = 40, t = 60, b = 50)
  )


```

```{r}
# Load libraries
library(dplyr)
library(tidyr)
library(plotly)
library(RColorBrewer)

# Step 1: Clean and reshape data
DATA_long <- DATA_sf %>%
  mutate(across(c(Developed_Open_Space,
                  Developed_Low_Intensity,
                  Developed_Medium_Intensity,
                  Developed_High_Intensity,
                  X_ENDC_RICHNESS_), as.numeric)) %>%
  pivot_longer(cols = starts_with("Developed_"),
               names_to = "Development_Type",
               values_to = "Percent_Cover") %>%
  filter(!is.na(Percent_Cover), Percent_Cover >= 0, Percent_Cover <= 100) %>%
  mutate(
    Development_Type = recode(Development_Type,
      "Developed_Open_Space" = "Open Space",
      "Developed_Low_Intensity" = "Low Intensity",
      "Developed_Medium_Intensity" = "Medium Intensity",
      "Developed_High_Intensity" = "High Intensity"
    )
  )

# Step 2: Define colors and create base plot
dev_levels <- unique(DATA_long$Development_Type)
colors <- brewer.pal(length(dev_levels), "Set2")
plot <- plot_ly()

# Step 3: Loop through each development type and add points + trend line
for (i in seq_along(dev_levels)) {
  dev_type <- dev_levels[i]
  df_sub <- DATA_long %>% filter(Development_Type == dev_type)
  
  # Fit linear model
  fit <- lm(X_ENDC_RICHNESS_ ~ Percent_Cover, data = df_sub)
  r2 <- round(summary(fit)$r.squared, 2)
  
  # Create predicted trend line
  line_df <- data.frame(
    Percent_Cover = seq(min(df_sub$Percent_Cover), max(df_sub$Percent_Cover), length.out = 100)
  )
  line_df$Predicted <- predict(fit, newdata = line_df)
  
  # Add scatter points
  plot <- plot %>%
    add_trace(
      data = df_sub,
      x = ~Percent_Cover + 0.1,
      y = ~X_ENDC_RICHNESS_,
      type = 'scatter',
      mode = 'markers',
      name = paste0(dev_type, " (R² = ", r2, ")"),
      marker = list(size = 5, opacity = 0.5, color = colors[i]),
      showlegend = TRUE
    ) %>%
    # Add bold regression line
    add_trace(
      data = line_df,
      x = ~Percent_Cover + 0.1,
      y = ~Predicted,
      type = 'scatter',
      mode = 'lines',
      line = list(color = colors[i], width = 3, dash = "solid", opacity = 1),
      name = paste0(dev_type, " Trend"),
      showlegend = FALSE
    )
}

# Step 4: Final layout
plot %>%
  layout(
    title = "Bee Species Richness vs. Land Development Intensity (with R²)",
    xaxis = list(title = "% Land Cover (log-scaled)", type = "log"),
    yaxis = list(title = "Bee Species Richness"),
    legend = list(title = list(text = "<b>Development Type</b>")),
    margin = list(l = 60, r = 40, t = 60, b = 50)
  )

```

