---
title: "California-BioClim-Project-PINC"
author: "Najeeba Shaikh"
date: "2025-06-12"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
getwd()
```

```{r}
setwd("/Users/najeebashaikh/Desktop/PINC_Climate_Change/Mini-Project 1 - Bees/")
```

```{r}
library("ggplot2")

library("chron")

library("RColorBrewer")

library("tidyr")

library("dplyr")

library("tidyverse")

library("mapview")

``` 

#read the dataset
```{r}
Cali_bioclim <- read.csv("california_bioclim.csv")
```

#Check the structure of your data using summary() and head()
```{r}
head(Cali_bioclim)
```

```{r}
summary(Cali_bioclim)
```

# Remove rows with missing values (NA)
```{r}

```

# Rename columns with their real-world names (e.g., BIO1 → Annual MeanTemp)
```{r}
BioClim_Cali <- Cali_bioclim %>%
  rename(Annual_Mean_Temperature = bio1, Mean_Diurnal_Range = bio2, Isothermality = bio3, Temperature_Seasonality = bio4, Max_Temperature_of_Warmest_Month = bio5, Min_Temperature_of_Warmest_Month = bio6, Temperature_Annual_Range = bio7, Mean_Temperature_of_Wettest_Quarter = bio8, Mean_Temperature_of_Driest_Quarter = bio9, Mean_Temperature_of_Warmest_Quarter = bio10, Mean_Temperature_of_Coldest_Quarter = bio11, Annual_Precipitation = bio12, Precipitation_of_Wettest_Month = bio13, Precipitation_of_Driest_Month = bio14, Precipitation_Seasonality = bio15, Precipitation_of_Wettest_Quarter = bio16, Precipitation_of_Driest_Quarter = bio17, Precipitation_of_Warmest_Quarter = bio18, Precipitation_of_Coldest_Quarter = bio19)
```

# Convert temperature columns from Celsius to Fahrenheit (original dataset)
# mutate(column_name_f = column_name_c * 9/5 + 32)
# bio1 - bio11 are temp
```{r}
Fahrenheit_data <- transform(Cali_bioclim,
                             bio1_F = bio1* 9/5 + 32, 
                             bio2_F = bio2* 9/5 + 32,
                             bio3_F = bio3* 9/5 + 32,
                             bio4_F = bio4 * 9/5 + 32,
                             bio5_F = bio5 * 9/5 + 32,
                             bio6_F = bio6 * 9/5 + 32,
                             bio7_F = bio7* 9/5 + 32,
                             bio8_F = bio8* 9/5 + 32,
                             bio9_F = bio9* 9/5 + 32,
                             bio10_F = bio10* 9/5 + 32,
                             bio11_F = bio11* 9/5 + 32)
```

# Use three mathematical functions like mean, sd
```{r}
mean(BioClim_Cali$Total_Specimens)
```

```{r}
sd(BioClim_Cali$Total_Specimens)
```

```{r}
abs(BioClim_Cali$Annual_Precipitation)
```

# Find and report the maximum species richness and temperature values
```{r}
max(BioClim_Cali$X_ENDC_RICHNESS_)
```

```{r}
max(BioClim_Cali$Mean_Temperature_of_Coldest_Quarter)
```

## Plot three different types of graphs
```{r}
ggplot(BioClim_Cali, aes(x = grid_cell, y = Annual_Mean_Temperature)) +
  geom_line(color = "skyblue", size = 1) +
  geom_area(fill = "lightblue", alpha = 0.4) +  # Fill under the line
  labs(
    title = "Annual Mean Temperature Across Grid Cells",
    x = "Grid Cell",
    y = "Annual Mean Temperature (°C)"
  ) +
  theme_minimal()
```
```{r}
ggplot(BioClim_Cali, aes(x = Precipitation_Seasonality, y = Annual_Precipitation)) +
  geom_hex(color = "pink", size = 1) +
  geom_area(fill = "yellow", alpha = 0.4) +  # Fill under the line
  labs(
    title = "Seasonal Precipitation vs Annual Precipitation",
    x = "Seasonal Precipitation",
    y = "Annual Precipitation"
  ) +
  theme_minimal()
```

```{r}
ggplot(BioClim_Cali, aes(x = Precipitation_of_Driest_Quarter, y = Precipitation_of_Warmest_Quarter, z = Precipitation_of_Coldest_Quarter), col = "Precipitation") +
  scale_color_brewer(palette = "Spectral") +
  geom_contour_filled(color = "forestgreen", size = 1) +
  geom_area(fill = "darkgreen", alpha = 0.2)
  
```
```{r}
ggplot(BioClim_Cali, aes(x = X_ENDC_RICHNESS_, y = Total_Specimens)) +
  geom_boxplot(fill = "skyblue", color = "darkblue", alpha = 0.7)
```


# Make a California map using the grid coordinates
```{r}
mapview(BioClim_Cali, xcol = "X_Axis_0_", ycol = "X_Axis_1_", crs = 3310, grid = TRUE)
```





